---
title: From prototype to production, choosing the right R and C++ tool in Rcpp
author: Thiago de Paula Oliveira
date:  "`r format(Sys.Date(), '%B %d, %Y')`"
slug: rcpp-start-with-r
categories:
  - R Programming
  - Data handling
tags:
  - C++
  - Rcpp
  - R
subtitle: ''
summary: 'what each tool solves, when to reach for it, and ready-to-paste code.'
authors: 
- admin
lastmod: '2025-08-06T09:15:54Z'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  html_document:
    keep_md: yes
    toc: true
---

<style>
/* ---------- CSS variables: change in one place ---------- */
:root{
  --font-main: 'Helvetica Neue',Arial,sans-serif;
  --font-mono: 'Courier New',Courier,monospace;
  --clr-bg:    #fafafa;
  --clr-text:  #333;
  --clr-brand: #2f6ab5;      /* darker → better contrast */
  --clr-brand-light: #3b80d1;
  --radius:    5px;
}

/* ---------- Page base ---------- */
body{
  font-family:var(--font-main);
  font-size:1rem;line-height:1.8;
  colour:var(--clr-text);
  background:var(--clr-bg);
  margin:0;padding:0 20px;
  text-align:justify;
}

/* ---------- Headings ---------- */
h1,h2,h3,h4,h5,h6{
  font-weight:600;
  margin:1.5em 0 .75em;
  line-height:1.2;
  colour:#0d0d0d;
}
h1{font-size:2rem;   border-bottom:3px solid var(--clr-brand);  padding-bottom:.3em;}
h2{font-size:1.75rem;border-bottom:2px solid var(--clr-brand);  padding-bottom:.3em;}
h3{font-size:1.5rem; colour:var(--clr-brand-light);            padding-bottom:.2em;}
h4{font-size:1.25rem;} h5{font-size:1.125rem;} h6{font-size:1rem;}

/* ---------- Links ---------- */
a{
  colour:var(--clr-brand-light);text-decoration:none;
  transition:colour .3s;
}
a:hover, a:focus-visible{colour:var(--clr-brand);text-decoration:underline;}
:focus-visible{outline:3px dashed var(--clr-brand);outline-offset:3px;}

/* ---------- Code Blocks ---------- */
pre,.code-input{
  background:#f5f5f5;border:1px solid #ddd;
  padding:10px;margin:20px 0;
  border-radius:var(--radius);overflow-x:auto;
  font-family:var(--font-mono);font-size:.9rem;
}
code{background:#f5f5f5;padding:2px 4px;border-radius:3px;font-family:var(--font-mono);}

/* ---------- Tables ---------- */
table{width:100%;border-collapse:collapse;margin-bottom:1.5em;text-align:left;}
th,td{padding:12px;border:1px solid #ddd;}
th{background:var(--clr-brand-light);colour:#fff;}

/* ---------- Highlight boxes ---------- */
.note-success,.note-warning{
  padding:10px;margin-bottom:1.5em;
  border-radius:var(--radius);colour:#000;
}
.note-success{background:#d6edd3;}
.note-warning{background:#cfbe7e;}

/* ---------- Images & CSS light-box ---------- */
img{
  display:block;max-width:100%;height:auto;
  border-radius:var(--radius);margin-bottom:1.5em;
  cursor:zoom-in;transition:transform .4s;
}
img:hover{transform:scale(1.02);}

/* container for the overlay created by JS */
#lightbox-overlay{
  position:fixed;inset:0;background:rgba(0,0,0,.85);
  display:flex;align-items:center;justify-content:center;
  z-index:1000;cursor:zoom-out;
}
#lightbox-overlay img{max-width:90%;max-height:90%;border-radius:var(--radius);}
</style>

<script>
/* minimal, pop-up-safe light-box */
document.addEventListener('DOMContentLoaded',()=>{
  document.querySelectorAll('img').forEach(img=>{
    img.addEventListener('click',()=>{
      /* build overlay only once */
      let ov=document.getElementById('lightbox-overlay');
      if(!ov){
        ov=document.createElement('div');
        ov.id='lightbox-overlay';
        ov.addEventListener('click',e=>e.currentTarget.remove());
        document.body.appendChild(ov);
      }
      ov.innerHTML=`<img src="${img.src}" alt="">`;
    });
  });
});
</script>

#  Prototype at light-speed with **inline**

`inline::cxxfunction()` compiles, links and returns an `R` function from a character string, ideal for quick sketches or teaching. 

```{r}
library(inline)
library(Rcpp)

src <- '
  // Declare two Rcpp numeric vectors, populated from the R arguments `a` and `b`
  Rcpp::NumericVector xa(a), xb(b);

  // Cache the lengths of those vectors in plain C++ ints for fast reuse
  int na = xa.size(), nb = xb.size();

  // Pre-allocate the output vector, with length (na + nb − 1) for a full 
  // discrete convolution
  Rcpp::NumericVector out(na + nb - 1);

  // Walk over every element of the first vector
  for (int i = 0; i < na; ++i)
    // Multiply the i-th element of `a` with every element of `b`
    for (int j = 0; j < nb; ++j)
        // Accumulate the product at the correct lag/index in the result
        out[i + j] += xa[i] * xb[j];

  // Return the filled result vector back to R
  return out;
'

conv <- cxxfunction(signature(a = "numeric", b = "numeric"),
                    body = src, plugin = "Rcpp")
conv(1:4, 2:5)
#> [1]  2  7 16 30 34 31 20
```

Choose **inline** when:

* you need an answer *now* without creating extra files;
* you are benchmarking or exploring an API shape;
* you want to inject ad-hoc `#include` blocks or compiler flags.

## Level-up with plugins and **Rcpp attributes**

### 4.1 Plugins

Plugins bolt extra headers and libraries onto an `inline` or attributes build—no manual `-I`/`-L` juggling.
The built-in `RcppArmadillo` plugin lets you write BLAS-backed linear algebra on the fly:

```{r}
library(RcppArmadillo)
src <- '
  // Armadillo + Rcpp headers are supplied automatically by the plugin
  using namespace Rcpp;

  /* ---- 1. Map the inputs -------------------------------------------------- */
  Rcpp::NumericVector  yr_(yr);            // wrap SEXP as NumericVector
  Rcpp::NumericMatrix  Xr_(Xr);            // wrap SEXP as NumericMatrix

  int n = Xr_.nrow();                      // rows  = #obs
  int k = Xr_.ncol();                      // cols  = #predictors

  // zero-copy Armadillo views
  arma::colvec y( yr_.begin(), yr_.size(), false );
  arma::mat    X( Xr_.begin(), n, k,       false );

  /* ---- 2. OLS -------------------------------------------------------------- */
  arma::colvec coef = arma::solve(X, y);
  arma::colvec res  = y - X * coef;
  double s2         = arma::dot(res, res) / (n - k);
  arma::colvec se   = arma::sqrt( s2 * arma::diagvec( arma::inv(X.t() * X) ) );

  /* ---- 3. Ship results back to R ------------------------------------------ */
  return List::create(_["coef"] = coef,
                      _["se"]   = se,
                      _["df"]   = n - k);
'

fastLm <- cxxfunction(signature(yr = "numeric", Xr = "numeric"),
                      body   = src,
                      plugin = "RcppArmadillo")

# Usage 
set.seed(123)

n <- 1000
k <- 3
X <- cbind(1, matrix(rnorm(n * k), n, k))  # 1st col = intercept
beta  <- c(5, 0.8, -1.2, 0.5)                 # true coefficients
y  <- X %*% beta + rnorm(n, sd = 2)           # synthetic data
res <- fastLm(y, X)
str(res)

fit <- lm(y ~ X[,2] + X[,3] + X[,4])  # lm() adds its own intercept

cbind(fastLm = res$coef,
      lm     = coef(fit))

microbenchmark::microbenchmark(
  fastLm = fastLm(y, X),
  lm     = { lm(y ~ X[,2] + X[,3] + X[,4]) },
  times  = 100
)
```


### 4.2 Rcpp Attributes

* Attributes internalise *inline*’s magic behind `[[Rcpp::export]]`.
* `sourceCpp()` compiles a file; 
* `cppFunction()` compiles from a string; 
* `evalCpp()` runs an expression, all while caching builds transparently.

```{r}
cpp_src <- paste(
  '#include <Rcpp.h>',
  'using namespace Rcpp;',
  '',
  '// [[Rcpp::export]]',
  'int fibonacci(int x) {',
  '  return (x < 2) ? x : fibonacci(x - 1) + fibonacci(x - 2);',
  '}',
  sep = "\n"
)
writeLines(cpp_src, "fibonacci.cpp")   # main copy
```

```{r}
Rcpp::sourceCpp("fibonacci.cpp")
fibonacci(20)     #> 6765
```

Use attributes for almost everything once you move beyond exploratory work.


## Quick decision guide

| Task                                              | Reach for …                                                                           |
| ------------------------------------------------- | ------------------------------------------------------------------------------------- |
| Package / production code                         | **Rcpp Attributes**                                                                   |
| One-off sketches, benchmarks, Stack Overflow post | **inline::cxxfunction()**                                                             |
| Extra C++ libraries (Armadillo, Eigen, GSL, ...)    | **Plugins** (built-in or custom)                                                      |
| Low-level debugging with `gdb`                    | Manual `R CMD SHLIB` + helper flags                                                   |
| Propagate C++ exceptions cleanly                  | Ensure `BEGIN_RCPP/END_RCPP` are present (added automatically by `inline` / Attributes) |

---

## Take-aways

* **Match R’s compiler** or live with hard-to-trace crashes.
* **`.Call()` via Rcpp** is the modern native interface—type, which is safe and header-driven.
* **inline** and **Attributes** shrink the compile–link–load loop to a single line.
* **Plugins** keep code portable while tapping heavy `C++` libraries.
* **BEGIN\_RCPP/END\_RCPP** keep your` C++` errors readable from R.


**Did you find this page helpful? Consider sharing it 🙌**