---
title: "Benchmarking Kendall's Tau in R and Rcpp"
date:  "`r format(Sys.Date(), '%B %d, %Y')`"
slug: "benchmarking-kendalls-tau-r-rcpp"
categories:
  - "R Programming"
  - "Statistical Computing"
tags:
  - "Rcpp"
  - "C++"
  - "Kendall's tau"
  - "rank correlation"
  - "microbenchmarking"
  - "performance"
subtitle: "R vs Rcpp with a fast inversion-count implementation"
summary: "Implement and benchmark a fast Kendall's tau-a in C++ via Rcpp against base R, discuss tie handling (tau-b), and when to move from R to C++."
authors:
  - admin
lastmod: "`r format(as.POSIXct(Sys.time(), tz = 'UTC'), '%Y-%m-%dT%H:%M:%SZ')`"
featured: false
image:
  caption: ""
  focal_point: ""
  preview_only: false
projects: []
output:
  blogdown::html_page:
    toc: true
    number_sections: true
    toc_depth: 3
---

# Overview

This post demonstrates how to move from a prototype in R to a compiled implementation with **Rcpp**, using **Kendall’s rank correlation** as a worked example. We implement a fast **tau-a** and **tau-b** estimator in `C++` (merge-sort inversion count), validate it against `stats::cor(..., method = "kendall")` on tie-free and non-tie-free data, and microbenchmark all approaches.

# Setup

```{r setup, results='hide', message=FALSE, warning=FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)

library(Rcpp)
library(microbenchmark)
library(dplyr)
library(ggplot2)
library(knitr)
set.seed(1)
```

# Kendall’s tau definitions

Let $\{(x_i,y_i)\}_{i=1}^n$ be paired observations, and consider all unordered index pairs $(i,j)$ with $i<j$. For each pair define

$$
  s_{ij} \,=\, \operatorname{sgn}(x_i-x_j)\,\operatorname{sgn}(y_i-y_j),
$$

with the convention $\operatorname{sgn}(0)=0$. Thus, $s_{ij}=+1$ for concordant, $-1$ for discordant, and $0$ if the pair is tied in $x$ or in $y$ (or both).

Classify pairs:

* **Concordant** if $(x_i-x_j)(y_i-y_j)>0$.
* **Discordant** if $(x_i-x_j)(y_i-y_j)<0$.
* **Tie in** $x$ if $x_i=x_j$ (regardless of $y$).
* **Tie in** $y$ if $y_i=y_j$ (regardless of $x$).

Define the counts

$$
  C = \sum_{1\le i<j\le n} \mathbf{1}\!\big((x_i-x_j)(y_i-y_j)>0\big),\quad
  D = \sum_{1\le i<j\le n} \mathbf{1}\!\big((x_i-x_j)(y_i-y_j)<0\big),
$$

and let $n_0=\binom{n}{2}$ be the total number of pairs.

If the $x$-values contain tie groups of sizes $\{t_{x,g}\}$ and the $y$-values tie groups $\{t_{y,h}\}$, the numbers of tied pairs (counting all ties, including pairs tied on both variables) are

$$
  T_x = \sum_g \binom{t_{x,g}}{2}, \qquad
  T_y = \sum_h \binom{t_{y,h}}{2}.
$$

Let $T_{xy}$ be the number of pairs tied **on both** variables. Then the calculating identities

$$
  n_0 = C + D + T_x + T_y - T_{xy}, \qquad \sum_{i<j} s_{ij} = C - D
$$

hold.

## Kendall’s $\tau_a$ (no tie correction)

$$
  \tau_a \,=\, \frac{C - D}{n_0} \,=\, \frac{1}{n_0}\sum_{i<j} s_{ij}.
$$

This coefficient lies in $[-1,1]$. It is appropriate when ties are absent or negligible; ties contribute $s_{ij}=0$ and therefore shrink $|\tau_a|$ toward 0.

## Kendall’s $\tau_b$ (tie-adjusted)

$$
  \tau_b \,=\, \frac{C - D}{\sqrt{\,(n_0 - T_x)\,(n_0 - T_y)\,}}.
$$

This rescales by the numbers of *comparable* pairs in each margin and preserves the $[-1,1]$ range. When there are no ties ($T_x=T_y=0$), we have $\tau_b=\tau_a$.

> **Convention on double ties.** Pairs tied on both variables are included in **both** $T_x$ and $T_y$, matching common software implementations of $\tau_b$. The separate term $T_{xy}$ is used only in the calculating identity above.

## Remarks and edge cases

It is important to keep in mind the following properties:

* **Degenerate margins.** If all $x$ values are equal (so $T_x=n_0$) or all $y$ values are equal ($T_y=n_0$), the denominator of $\tau_b$ is zero and $\tau_b$ is undefined (software often returns `NaN`). In these cases, $\tau_a=0$.
* **Invariance.** Both $\tau_a$ and $\tau_b$ are invariant to strictly monotone transformations of either variable.
* **No ties.** If there are no ties, then $C+D=n_0$ and both measures coincide: $\tau=\frac{C-D}{n_0}$.

# Implementation in R (Rcpp)

With the definitions in place, we can move from notation to a compact implementation. To enumerate all $n_0=\binom{n}{2}$ pairs efficiently, we’ll use a small `C++` routine via **Rcpp**. The function below computes Kendall’s $\tau_a$ by tallying concordant and discordant pairs and returning $(C-D)/n_0$. When ties are common, $\tau_b$ is usually preferable; we present both approaches, but the reason to have $\tau_a$ here because it is simple and fast, and it coincides with $\tau_b$ when there are no ties. 

## Rcpp inline implementation ($\tau_a$)

The C++ function below is intended to be compiled inline from `R` and called as a regular `R` function. (A short note on validation follows after the code.)


```{r rcpp_inline, warning = FALSE, message=FALSE, cache=FALSE}
Rcpp::cppFunction(code = '
#include <Rcpp.h>
#include <algorithm>
#include <numeric>
#include <cmath>
using namespace Rcpp;

// Merge step that counts inversions while sorting.
long long merge_count(IntegerVector& arr, IntegerVector& temp,
                      int left, int mid, int right) {
  int i = left, j = mid, k = left;
  long long inv_count = 0;

  while (i < mid && j <= right) {
    if (arr[i] <= arr[j]) {
      temp[k++] = arr[i++];
    } else {
      temp[k++] = arr[j++];
      inv_count += (mid - i); // elements i..mid-1 are > arr[j-1]
    }
  }
  while (i < mid)    temp[k++] = arr[i++];
  while (j <= right) temp[k++] = arr[j++];

  for (int idx = left; idx <= right; ++idx) arr[idx] = temp[idx];
  return inv_count;
}

// Recursive sort that returns inversion count.
long long sort_count(IntegerVector& arr, IntegerVector& temp, int left, int right) {
  if (right - left < 1) return 0LL;
  int mid = left + (right - left) / 2;
  long long inv = sort_count(arr, temp, left, mid);
  inv += sort_count(arr, temp, mid + 1, right);
  inv += merge_count(arr, temp, left, mid + 1, right);
  return inv;
}

// [[Rcpp::export]]
double kendall_tau_a_cpp(NumericVector x, NumericVector y) {
  int n = x.size();
  if (n < 2) return NA_REAL;

  // Indices that sort x.
  IntegerVector idx(n);
  std::iota(idx.begin(), idx.end(), 0);
  std::sort(idx.begin(), idx.end(),
            [&x](int i, int j){ return x[i] < x[j]; });

  // y reordered by sorted x; discretise for stable comparison.
  IntegerVector y_ord(n);
  for (int i = 0; i < n; ++i) {
    y_ord[i] = (int) std::floor(y[idx[i]] * 1e8);
  }

  // Count discordances as inversions in y_ord.
  IntegerVector temp(n);
  long long discord = sort_count(y_ord, temp, 0, n - 1);

  // tau-a.
  double n0 = double(n) * double(n - 1) / 2.0;
  return (n0 - 2.0 * discord) / n0;
}
', rebuild = TRUE)
```

## Rcpp inline implementation (tau-b)

This `C++` routine computes Kendall’s **tau-b** (tie-adjusted). It first orders observations by $x$ (breaking ties by $y$), then uses a Fenwick tree to accumulate $S=C-D$ by comparing each group (equal $x$) only to previously seen values (so pairs tied on either variable do not enter the numerator). The tie counts $T_x$ and $T_y$ are computed explicitly and used in the denominator $\sqrt{(n_0-T_x)(n_0-T_y)}$, yielding a coefficient in $[-1,1]$ that matches `stats::cor(..., method = "kendall")` on tied data. Time complexity is $O(n\log n)$. Floating inputs are discretised (scale factor $10^8$) to stabilise equality checks; for heavily discretised data we may prefer pre-ranking to integers.

```{r rcpp_inline_taub, warning = FALSE, message=FALSE, cache=FALSE}
Rcpp::cppFunction(code = "
#include <Rcpp.h>
#include <algorithm>
#include <numeric>
#include <cmath>
#include <vector>
using namespace Rcpp;

struct Fenwick {
  std::vector<long long> t;
  int n;
  Fenwick(int n) : t(n + 1, 0), n(n) {}
  void add(int i, long long v){
    for(; i <= n; i += i & -i) t[i] += v;
  }
  long long sum(int i) const {
    long long s = 0;
    for(; i > 0; i -= i & -i) s += t[i];
    return s;
  }
};  // 

// [[Rcpp::export]]
double kendall_tau_b_cpp(NumericVector x, NumericVector y) {
  const int n = x.size();
  if (n < 2) return NA_REAL;

  // Discretise to integers for stable tie detetion with floating inputs.
  std::vector<long long> xi(n), yi(n);
  for (int i = 0; i < n; ++i) {
    xi[i] = (long long) std::floor(x[i] * 1e8);
    yi[i] = (long long) std::floor(y[i] * 1e8);
  }

  // Order by x then y
  std::vector<int> ord(n);
  std::iota(ord.begin(), ord.end(), 0);
  std::sort(ord.begin(), ord.end(), [&](int a, int b){
    if (xi[a] != xi[b]) return xi[a] < xi[b];
    return yi[a] < yi[b];
  });

  // Create T_x (pairs tied on x)
  long long T_x = 0;
  for (int i = 0; i < n; ){
    int j = i + 1;
    while (j < n && xi[ord[j]] == xi[ord[i]]) ++j;
    long long g = j - i;
    T_x += g * (g - 1) / 2;
    i = j;
  }

  // Create T_y (pairs tied on y)
  std::vector<long long> y_sorted = yi;
  std::sort(y_sorted.begin(), y_sorted.end());
  long long T_y = 0;
  for (int i = 0; i < n; ){
    int j = i + 1;
    while (j < n && y_sorted[j] == y_sorted[i]) ++j;
    long long g = j - i;
    T_y += g * (g - 1) / 2;
    i = j;
  }

  // Coordinate-compress y
  std::vector<long long> y_unique = y_sorted;
  y_unique.erase(std::unique(y_unique.begin(), y_unique.end()), y_unique.end());
  auto rank_of = [&](long long v){
    return (int)(std::lower_bound(y_unique.begin(), y_unique.end(), v) - 
                                  y_unique.begin()) + 1; // 1-based
  };

  // S = C - D via BIT; group by equal x so x-ties contribute zero
  Fenwick bit((int) y_unique.size());
  long long processed = 0;
  long long S = 0;

  for (int i = 0; i < n; ){
    int j = i + 1;
    while (j < n && xi[ord[j]] == xi[ord[i]]) ++j;

    // Compare current x-group against all prior items (strict < and >)
    for (int k = i; k < j; ++k) {
      int idx = ord[k];
      int r = rank_of(yi[idx]);
      // y_prev < y
      long long less    = bit.sum(r - 1);   
      // y_prev <= y
      long long leq     = bit.sum(r);       
      // y_prev > y
      long long greater = processed - leq;  
      // +1 for concordant, -1 for discordant
      S += (less - greater);
    }

    // Insert current group into the tree
    for (int k = i; k < j; ++k) {
      int r = rank_of(yi[ord[k]]);
      bit.add(r, 1);
    }
    processed += (j - i);
    i = j;
  }

  const double n0      = (double) n * (double) (n - 1) / 2.0;
  const double denom_x = n0 - (double) T_x;
  const double denom_y = n0 - (double) T_y;
  if (denom_x <= 0.0 || denom_y <= 0.0) return NA_REAL;

  const double denom = std::sqrt(denom_x * denom_y);
  return (double) S / denom;
}
")
```

# Sanity checks considering tie-free vs tie-heavy data

In practice, Kendall’s rank correlation behaves differently depending on the prevalence of ties. The **tau-a** normalisation uses the total number of unordered pairs $n_0=\binom{n}{2}$ and does not adjust for ties, whereas **tau-b** rescales by $\sqrt{(n_0-T_x)(n_0-T_y)}$ to remove the influence of ties in $x$ and $y$. Base R’s `cor(..., method = "kendall")` implements a tie-aware estimator (tau-b–type normalisation). The aim of this section is therefore twofold (i) to show agreement amongst implementations when ties are essentially absent, and (ii) to illustrate the expected divergence of tau-a from tie-aware estimators when ties are pervasive, typical of discretised scores or coarse measurements in breeding datasets.


## Tie-free (continuous) data

With continuous, independent draws (here, Gaussian), ties are virtually absent. Under this regime, tau-a and tau-b coincide up to Monte Carlo error, and both agree with base `R`. Any residual differences are due to finite-sample variability and numerical details (e.g., discretisation safeguards in the `C++` code). This scenario serves as a basic correctness check before considering tie handling.


```{r sanity_noties}
set.seed(1)
n_noties <- 5000L
rho   <- 0.6
Sigma <- matrix(c(1, rho, rho, 1), 2, 2)

Z <- matrix(rnorm(2L * n_noties), ncol = 2) 
L <- chol(Sigma)
XY <- Z %*% L

x0 <- XY[, 1]
y0 <- XY[, 2]

tau_a_noties  <- kendall_tau_a_cpp(x0, y0)
tau_b_noties  <- kendall_tau_b_cpp(x0, y0)
tau_r_noties  <- suppressWarnings(stats::cor(x0, y0, method = "kendall"))

dplyr::tibble(
  scenario   = "no ties",
  tau_a_cpp  = tau_a_noties,
  tau_b_cpp  = tau_b_noties,
  tau_baseR  = tau_r_noties,
  abs_diff_a_vs_base = abs(tau_a_noties - tau_r_noties),
  abs_diff_b_vs_base = abs(tau_b_noties - tau_r_noties)
) |>
  knitr::kable(digits = 6, caption = "Tie-free data: Kendall estimates (tau-a, tau-b, base R)")
```
Note that for a **bivariate normal** with Pearson correlation $\rho$, the population Kendall’s tau is

$$
\boxed{\;\tau \;=\; \frac{2}{\pi}\,\arcsin(\rho)\;}
$$

So with $\rho=0.6$,

$$
\tau_{\text{theory}}
= \frac{2}{\pi}\arcsin(0.6)
\approx 0.409.
$$

## Tie-heavy data we have $\tau_b \approx$ base R, $\tau_a$ differs

When many observations share identical values (here enforced by sampling small integers), a substantial fraction of pairs are tied on $x$, on $y$, or on both. Tau-b explicitly discounts such tied pairs in the denominator and therefore remains on the same scale as base `R`. In contrast, tau-a retains $n_0$ in the denominator while excluding tied pairs from the numerator, pulling estimates towards zero as ties increase. In applied settings (e.g., ordinal disease scores, rounded phenotypes, or coarsened environmental indices), a tie-adjusted coefficient such as tau-b is generally preferable.

```{r sanity_ties}
set.seed(2)
n_ties <- 5000L
# Create many ties by sampling integers (stronger ties than rounding)
x1 <- sample.int(40L, n_ties, replace = TRUE)
y1 <- sample.int(50L, n_ties, replace = TRUE)

tau_a_ties  <- kendall_tau_a_cpp(x1, y1)
tau_b_ties  <- kendall_tau_b_cpp(x1, y1)
tau_r_ties  <- suppressWarnings(stats::cor(x1, y1, method = "kendall"))

dplyr::tibble(
  scenario   = "many ties",
  tau_a_cpp  = tau_a_ties,
  tau_b_cpp  = tau_b_ties,
  tau_baseR  = tau_r_ties,
  abs_diff_a_vs_base = abs(tau_a_ties - tau_r_ties),
  abs_diff_b_vs_base = abs(tau_b_ties - tau_r_ties)
) |>
  knitr::kable(digits = 6, 
               caption = "Tie-heavy data: Kendall estimates (tau-a, tau-b, base R)")
```


# Benchmark design & microbenchmarks (no ties vs many ties)

We benchmark three implementations on $n=10{,}000$ pairs. A `C++` **tau-a** (merge-sort inversion count), a `C++` **tau-b** (Fenwick tree with explicit tie counts), and base `R`’s Kendall (tie-aware, $O(n\log n)$). Each method is run 50 times after a warm-up to amortise compilation and jit effects. Results are summarised by the median and IQR of wall-clock times. Absolute timings will depend on hardware and compilation flags. Note that the principal comparison is relative performance under differing tie structures.

```{r bench_setup, message=FALSE}
set.seed(3)
n_bench <- 10000L

# No-ties benchmark data (here we use a simple independent samples)
xb <- rnorm(n_bench)
yb <- rnorm(n_bench)

# Many-ties benchmark data
xbt <- sample.int(200L, n_bench, replace = TRUE)
ybt <- sample.int(220L, n_bench, replace = TRUE)

# Warm-up (compile and JIT call paths)
invisible(kendall_tau_a_cpp(xb,  yb))
invisible(kendall_tau_b_cpp(xbt, ybt))
invisible(stats::cor(xb,  yb,  method = "kendall"))
invisible(stats::cor(xbt, ybt, method = "kendall"))
```

### Microbenchmark: tie-free data

On essentially tie-free inputs, all algorithms are $O(n\log n)$ but differ in constants. The tau-a routine often has a slight advantage because it avoids tie searching. Agreement in estimates alongside stable timing indicates the implementations are behaving as expected in the, let's say, ``easy" regime.

```{r bench_noties, warning=FALSE}
mb_noties <- microbenchmark::microbenchmark(
  cpp_tau_a  = kendall_tau_a_cpp(xb, yb),
  cpp_tau_b  = kendall_tau_b_cpp(xb, yb),
  base_kendall = stats::cor(xb, yb, method = "kendall"),
  times = 50L
)

mb_noties_sum <- as.data.frame(mb_noties) |>
  dplyr::group_by(expr) |>
  dplyr::summarise(
    median_ms = median(time) / 1e6,
    iqr_ms    = IQR(time) / 1e6,
    min_ms    = min(time) / 1e6,
    max_ms    = max(time) / 1e6,
    .groups = "drop"
  ) |>
  dplyr::arrange(median_ms)

knitr::kable(mb_noties_sum, digits = 2,
             caption = "Microbenchmark (no ties, n = 10,000): tau-a (C++), tau-b (C++), base R")
```

```{r plot_bench_noties, fig.width=7, fig.height=4}
ggplot2::ggplot(as.data.frame(mb_noties),
                ggplot2::aes(x = expr, y = time/1e6)) +
  ggplot2::geom_boxplot(outlier.alpha = 0.15) +
  ggplot2::labs(x = NULL, y = "Time (ms)",
       title = "No ties: kendall_tau_a_cpp / kendall_tau_b_cpp / base R") +
  ggplot2::theme_minimal(base_size = 12)
```

### Microbenchmark: many ties

With many ties, tie-aware methods incur modest overhead for counting and handling groups, yet remain $O(n\log n)$. The tau-a routine may be fastest but is estimating a different quantity in this regime; its computational advantage should not be traded off against interpretability when ties matter. In most breeding or agricultural datasets with discrete or rounded measures, the tau-b (or base `R`) path is the appropriate comparator.

```{r bench_ties, warning=FALSE}
mb_ties <- microbenchmark::microbenchmark(
  cpp_tau_a    = kendall_tau_a_cpp(xbt, ybt),
  cpp_tau_b    = kendall_tau_b_cpp(xbt, ybt),
  base_kendall = stats::cor(xbt, ybt, method = "kendall"),
  times = 50L
)

mb_ties_sum <- as.data.frame(mb_ties) |>
  dplyr::group_by(expr) |>
  dplyr::summarise(
    median_ms = median(time) / 1e6,
    iqr_ms    = IQR(time) / 1e6,
    min_ms    = min(time) / 1e6,
    max_ms    = max(time) / 1e6,
    .groups = "drop"
  ) |>
  dplyr::arrange(median_ms)

knitr::kable(mb_ties_sum, digits = 2,
             caption = "Microbenchmark (many ties, n = 10,000): tau-a (C++), tau-b (C++), base R")
```

```{r plot_bench_ties, fig.width=7, fig.height=4}
ggplot2::ggplot(as.data.frame(mb_ties),
                ggplot2::aes(x = expr, y = time/1e6)) +
  ggplot2::geom_boxplot(outlier.alpha = 0.15) +
  ggplot2::labs(x = NULL, y = "Time (ms)",
       title = "Many ties: kendall_tau_a_cpp / kendall_tau_b_cpp / base R") +
  ggplot2::theme_minimal(base_size = 12)
```
As you probably noted until here, in both scenarios $(n=10{,}000)$, the `C++` implementations are orders of magnitude faster than base R. Specifically in the last scenario:

* **cpp tau-a** has a median $1.21$ ms being $\approx 1,269 \times$ faster than base `R` ($1,535.89$ ms).
* **cpp tau-b:** has median $2.38$ ms being $\approx 645 \times$ faster than base `R`.
* **tau-b vs tau-a:** tau-b is $\approx 2.0 \times$ slower than tau-a, reflecting the extra work needed for tie handling.

The IQRs (0.18 ms for tau-a; 0.46 ms for tau-b; 214.89 ms for base `R`) indicate the `C++` timings are not only faster but also far more stable. Practically, this means we can use the tie-adjusted **tau-b C++** routine without sacrificing performance; the $\approx 2.0 \times$ overhead relative to tau-a is negligible in absolute terms, while the gap to base `R` remains several orders of magnitude.

# Reproducibility

```{r session}
sessionInfo()
```

**Did you find this page helpful? Consider sharing it 🙌**

# References

Kendall, M. G., & Gibbons, J. D. (1990). \emph{Rank Correlation Methods} (5th ed.). Oxford University Press.

Agresti, A. (2010). \emph{Analysis of Ordinal Categorical Data} (2nd ed.). Wiley. <https://doi.org/10.1002/9780470594001>

# How to cite this post

Oliveira, T. de paula. (2025, August 9). Benchmarking Kendall’s tau in R and Rcpp.
<https://prof-thiagooliveira.netlify.app/post/benchmarking-kendalls-tau-r-rcpp/>


<style>
/* ====== Post-only layout + typography polish (Hugo Blox / Tailwind) ====== */
/* Place this at the very end of the post so it wins the cascade. */

/* 0) Widen the page shell Hugo Blox uses around articles */
.page-body > .mx-auto,
.page-body .max-w-screen-xl{
  max-width: 100vw !important;
  width: 100% !important;
  padding-left: 0 !important;
  padding-right: 0 !important;
}

/* 1) Remove the Tailwind max-w cap on the <main> inside <article> */
.page-body article > main{
  max-width: none !important;  /* beats .max-w-6xl */
  width: 100% !important;
}

/* 2) Control the actual reading width (fluid per breakpoint) */
.page-body article .prose{
  /* Base: a touch larger with comfy line height */
  font-size: clamp(1rem, 0.96rem + 0.25vw, 1.12rem);
  line-height: 1.75;
  text-align: left;
  margin-inline: auto;

  /* Reading width: scale up on big screens, but keep lines readable */
  max-width: 86ch !important; /* default desktop */
}
@media (min-width: 1024px){  /* lg */
  .page-body article .prose{ max-width: 96ch !important; }
}
@media (min-width: 1280px){  /* xl */
  .page-body article .prose{ max-width: 102ch !important; }
}
@media (min-width: 1536px){  /* 2xl / very wide */
  .page-body article .prose{ max-width: 108ch !important; }
}

/* 3) Phones/tablets: full width with side padding */
@media (max-width: 768px){
  .page-body article .prose{
    max-width: 100% !important;
    padding-inline: 1rem;
  }
}

/* 4) Give the article more room by slimming sidebars on wide screens */
@media (min-width: 1280px){
  .hb-sidebar-container, .hb-toc { width: 12rem !important; } /* was 16rem */
}
@media (max-width: 1279.98px){
  .hb-sidebar-container{ display:none !important; } /* hide sidebar under xl */
}

/* --------- Clean, professional type polish (scoped to post content) -------- */
.page-body article .prose p{
  margin: 0 0 1.15em;
  text-wrap: pretty;
  hyphens: auto;
}

.page-body article .prose h1{
  font-size: clamp(1.9rem, 1.6rem + 1.2vw, 2.3rem);
  margin: 1.2em 0 .5em;
  padding-bottom: .25em;
  border-bottom: 2px solid #e5e7eb;
}
.page-body article .prose h2{
  font-size: clamp(1.4rem, 1.2rem + 0.6vw, 1.7rem);
  margin: 1.35em 0 .4em;
  padding-bottom: .2em;
  border-bottom: 1px solid #e5e7eb;
}
.page-body article .prose h3{
  font-size: clamp(1.15rem, 1.05rem + 0.35vw, 1.35rem);
  margin: 1.1em 0 .3em;
}

/* Links: subtle underline-on-hover */
.page-body article .prose a{
  color: #2f6ab5;
  text-decoration: none;
  border-bottom: 1px solid rgba(47,106,181,.25);
}
.page-body article .prose a:hover{
  color: #1f4f8f;
  border-bottom-color: currentColor;
}

/* Code blocks & inline code */
.page-body article .prose pre{
  background: #f6f8fa;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  padding: 12px 14px;
  overflow: auto;
}
.page-body article .prose code{
  background: #f6f8fa;
  border: 1px solid #e5e7eb;
  border-radius: 5px;
  padding: .15em .35em;
  font-size: .95em;
}
.page-body article .prose pre code{
  background: none; border: 0; padding: 0; font-size: 0.95em;
}

/* Tables */
.page-body article .prose table{
  width: 100%;
  border-collapse: collapse;
  margin: 1.2rem 0;
  font-variant-numeric: tabular-nums;
}
.page-body article .prose th,
.page-body article .prose td{
  border: 1px solid #e5e7eb;
  padding: .6rem .75rem;
}
.page-body article .prose thead th{
  background: #2f6ab5;
  color: #fff;
  text-align: left;
}

/* Images & figures */
.page-body article .prose img{ border-radius: 6px; }

/* Optional: allow “full-bleed” wide elements
   Add class="wide" to a block (table/pre/img wrapper) to span the viewport. */
.page-body article .prose .wide{
  width: 100vw;
  position: relative;
  left: 50%;
  right: 50%;
  margin-left: -50vw;
  margin-right: -50vw;
  padding-inline: clamp(12px, 4vw, 36px);
}

/* Footer timestamp spacing */
.page-body article time{ margin-top: 2rem; display: block; }

/* Dark mode tweaks */
html.dark .page-body article .prose pre,
html.dark .page-body article .prose code{
  background: #111826;
  border-color: #253041;
}
html.dark .page-body article .prose thead th{ background: #5aa0ff; }
</style>

