---
title: "Courses"
summary: "Education and training" 
author: "Thiago de Paula Oliveira"
date: "`r Sys.Date()`"
weight: 10
highlight: false
type: book
output:
  blogdown::html_page:
    number_sections: false
    toc: false
    css: "my_style.css"
---

```{r setup, include=FALSE}
# Configuring global parameters
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error=FALSE, warning = FALSE,
                      fig.align = "center", fig.height = 6, fig.width = 6,
                      class.source = "lineAnchors")
```
<hr>
## Summary

```{r, echo=FALSE}
type <- as.factor(c("Genetics", "Programming language", "Statistical modelling and correlation", "Describing and displaying data", 
          "Hypothesis tests and confidence intervals", "Workflow"))
y <- c(1, # Genetics
       6, # Programming language
      12, # Statistical modelling and correlation
       6, # Describing and displaying data
       1, # Hypothesis tests and confidence intervals
       2  # Workflow
      )
library(tidyverse)
tibble(type, y) %>%
  ggplot(aes(y=y, fill=type, x = "")) + 
  geom_col(color = "black") +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  geom_text(aes(label = y),
            position = position_stack(vjust = 0.5)) +
  labs(fill = "Type") +
  theme_void() 
```


<hr>

## 2021

1. Workflows with Nextflow, University of Edinburgh
2. [Introduction to Bash Shell Scripting](https://www.coursera.org/account/accomplishments/certificate/3FXUKF7J65G8), Coursera Project Network
3. Genome-wide prediction of complex traits in humans, plants and animals
4. UKRI-BBSRC Workshop on Computing in the Biosciences
5. World Meeting of the International Society for Bayesian Analysis

<br>
<hr>
## 2020

1. [Programming Fundamentals](https://www.coursera.org/account/accomplishments/certificate/HW36TQ7CA337), Coursera, Duke University, USA.

<br>
<hr>
## 2019

1. Survival Analysis in R. DataCamp, USA.
2. Building Web Applications in R with Shiny: Case Studies Course. DataCamp, USA.
3. Building Dashboards with shinydashboard. DataCamp, USA.
4. Building Web Applications in R with Shiny. DataCamp, USA.
5. Introduction to Python. DataCamp, USA.
6. Statistical Modeling in R (Part 1). DataCamp, USA
7. Intermediate R. DataCamp, USA
8. Introduction to R. DataCamp, USA

<br>
<hr>
## 2018 

1. Machine Learning Toolbox. DataCamp, USA

<br>
<hr>
## 2016
1. Longitudinal and Incomplete Data – USP

<br>
<hr>
## 2015
1. Regression Models – Coursera, MOOC, USA
2. Short curse on Dimensionality Reduction – USP
3. Additive Generalized Models with P-splines – RBras
4. Exploring interactive graphical interfaces in R – RBras
5. Exploring the Flexibility of Linear Mixed Models – RBras
6. Special Topics in Multivariate Analysis – RBras

<br>
<hr>
## 2014

1. Generalized Additive Models with P-splines – USP

<br>
<hr>
## 2013

1. [Statistics: Making Sense of Data](https://www.coursera.org/api/legacyCertificates.v1/spark/statementOfAccomplishment/970433~3060810/pdf) – Coursera, MOOC, USA
2. [Mathematical Biostatistics Boot Camp](https://www.coursera.org/api/legacyCertificates.v1/spark/statementOfAccomplishment/970405~3060810/pdf) – Coursera, MOOC, USA
3. Introduction to Categorical Data Analysis – USP
4. Structural Equations Models – USP
5. Some Important Topics of Asymptotic Theory – USP